---
title: "Annotation and Gene Set Walkthrough"
author: "Mark Dunning"
date: "20 February 2020"
output: html_notebook
---

Load the data from the file

```{r message=FALSE, warning=FALSE}
library(DESeq2)
library(dplyr)
de_status <- readRDS("Robjects/de_status.rds")
```

Convert to a data frame and order by p-value

```{r}
results_ordered <- as.data.frame(results(de_status)) %>% 
  tibble::rownames_to_column("SYMBOL") %>% 
  arrange(padj)
```

Check the table...

```{r}
head(results_ordered)
```

Use the annotation package to get pathway IDs

```{r}
# to get names of keys
# keytypes(org.Hs.eg.db)

# to see what you can retrieve
# columns(org.Hs.eg.db)
library(org.Hs.eg.db)
anno_GO <- AnnotationDbi::select(org.Hs.eg.db,
                      keytype = "SYMBOL",
                      keys = results_ordered$SYMBOL[1:10],
                      columns = "GO")
head(anno_GO)
```

How many times do we see particular ontology IDs?

```{r}
count(anno_GO,GO)
```

It seems that *GO:0000981* comes up a lot. Is this significant? We need to know how many genes belong to the pathway.

```{r}
pathway_genes <- AnnotationDbi::select(org.Hs.eg.db,
                      keytype = "GO",
                      keys="GO:0000981",
                      columns = "SYMBOL") %>% 
  filter(!duplicated(SYMBOL)) %>% 
  pull(SYMBOL)

length(pathway_genes)
```

To assess the significance we will add extra columns to the table; is each gene significant? does the gene belong to our pathway of interest?

```{r}
results_GO <- results_ordered %>% mutate(is_DE = padj < 0.05) %>% 
  mutate(in_pathway = SYMBOL %in% pathway_genes)
results_GO
```

A chi-squared test can be used to assess significance; this is the basis for gene-set testing

```{r}
chisq.test(table(pathway=results_GO$in_pathway, DE=results_GO$is_DE))

```

However, this is just for one pathway. To test *all* pathways there are various web tools we can use. The tools requires text files that we can create with R.

*results/genes_for_GO.txt* is a file containing names of all DE genes.

```{r}
results_GO %>% 
  filter(is_DE) %>% 
  pull(SYMBOL) %>% 
  write.table("results/genes_for_GO.txt",row.names = FALSE,quote=FALSE)
```

*genes_universe.txt* is all genes that we tested.

```{r}
results_GO %>% 
  pull(SYMBOL) %>% 
  write.table("results/gene_universe.txt",row.names=FALSE,quote=FALSE)
```

these two files can be uploaded to [GOrilla](http://cbl-gorilla.cs.technion.ac.il/) for example

## GSEA

An alternative to the gene-set testing approach above is GSEA first proposed by the Broad institute. In this method, we don't require a hard cut-off for significance. However, we have to order (`arrange`) our table by test statistic.

A java application for GSEA is available [online](https://www.gsea-msigdb.org/gsea/downloads.jsp). The input for this tool is a file with two-columns; gene name and test statistic.

The GSEA tool seems to require the input file to have the file extension `.rnk`. We also have to remove any `NA` values.


```{r}

results_ordered %>% 
  filter(!duplicated(SYMBOL)) %>% 
  filter(!is.na(stat)) %>% 
  arrange(stat) %>% dplyr::select(SYMBOL, stat) %>% 
  write.table("results/gsea_input.rnk",sep="\t",quote=FALSE,row.names=FALSE)

```
